---
title: "01_Data processing"
authors: "Trudy, Jenny, Janice, ZZ"
output: html_document
date: "2026-04-04"
---

```{r setup}

library(readxl)
library(dplyr)
library(lubridate)
library(tidyr)
library(forecast)
library(tseries)
library(ggplot2)

```

We use three raw data series: oil price, U.S. oil stock, and U.S. domestic oil production. Oil price is the dependent variable of interest. U.S. oil stock and U.S. domestic oil production are included to account for supply-side conditions that may influence oil prices. We hypothesize that geopolitical events affect oil prices indirectly by changing U.S. oil inventories and, in turn, domestic oil production. These changes may then contribute to movements in oil prices.

## Data Input

```{r}
wti_raw <- read_excel("Data/Cushing, OK WTI Spot Price FOB (Dollars per Barrel).xls")
stock_raw <- read_excel("Data/U.S. Ending Stocks of Crude Oil (Thousand Barrels).xls")
prod_raw <- read_excel("Data/U.S. Field Production of Crude Oil (Thousand Barrels per Day).xls")
```

We are looking at a 10-year range because this would be long enough to observe how geopolitical events influenced oil price in the previous years. With that said, we will only select data

## Data Wrangling

```{r}
wti_df <- data.frame(
  Date  = as.Date(wti_raw[[1]], format = "%Y/%m/%d"),
  Price = as.numeric(wti_raw[[2]])
) %>%
  na.omit() %>%
  arrange(Date)

wti_ts <- ts(wti_df[,2], start = c(1986,1), frequency = 12)

wti_sub <- window(wti_ts, start = c(2005, 1), end = c(2025, 12))

# Plot WTI training series
plot(wti_sub, main = "WTI Price (2000-2025)", ylab = "Price", xlab = "Time")

# Stationarity check
adf.test(wti_sub)

fit_arima <- auto.arima(wti_sub)
fit_arima

```

# Involve fundamentals (production and inventores) and shock (war dummy)
```{r}
stock_df <- data.frame(
  Date  = as.Date(stock_raw[[1]], format = "%Y/%m/%d"),
  Price = as.numeric(stock_raw[[2]])
) %>%
  na.omit() %>%
  arrange(Date)

prod_df <- data.frame(
  Date  = as.Date(prod_raw[[1]], format = "%Y/%m/%d"),
  Price = as.numeric(prod_raw[[2]])
) %>%
  na.omit() %>%
  arrange(Date)

prod_ts  <- ts(prod_df[,2],  start = c(1920, 1), frequency = 12)
stock_ts <- ts(stock_df[,2], start = c(2005, 1), frequency = 12)

# Subset to same training period (2000–2025)
prod_sub  <- window(prod_ts,  start = c(2005,1), end = c(2025,12))
stock_sub <- window(stock_ts, start = c(2005,1), end = c(2025,12))

time_index <- time(wti_sub)

war_dummy <- ifelse(
  time_index >= 2022.1 & time_index < 2022.4,
  1, 0
)

xreg_all <- cbind(prod_sub, stock_sub, war_dummy)

fit_arimax <- auto.arima(wti_sub, xreg = xreg_all)
fit_arimax
```

# Use Gulf War to train and test for Ukraine War
```{r}
# Gulf War window
gulf_train <- window(wti_ts, start = c(1986, 1), end = c(1990, 6))
gulf_test  <- window(wti_ts, start = c(1990, 7), end = c(1991, 1))

#log transform
gulf_train_log <- log(gulf_train)
gulf_test_log  <- log(gulf_test)

#Gulf War baseline model
fit_gulf_arima <- auto.arima(gulf_train_log)
fit_gulf_ets   <- ets(gulf_train_log)
fit_gulf_tbats <- tbats(gulf_train_log)

#Forecast Gulf War based on baseline
h_gulf <- length(gulf_test_log)

fc_gulf_arima <- forecast(fit_gulf_arima, h = h_gulf)
fc_gulf_ets   <- forecast(fit_gulf_ets, h = h_gulf)
fc_gulf_tbats <- forecast(fit_gulf_tbats, h = h_gulf)

# War shock effects
shock_gulf_arima <- as.numeric(gulf_test_log - fc_gulf_arima$mean)
shock_gulf_ets   <- as.numeric(gulf_test_log - fc_gulf_ets$mean)
shock_gulf_tbats <- as.numeric(gulf_test_log - fc_gulf_tbats$mean)
```

# Ukraine War's effect forecast
```{r}
# Ukraine window（same length as Gulf War）
ukr_train <- window(wti_ts, start = c(2018, 1), end = c(2022, 1))
ukr_test  <- window(wti_ts, start = c(2022, 2), end = c(2022, 8))  # 7 months

#log transform
ukr_train_log <- log(ukr_train) 
ukr_test_log  <- log(ukr_test)

#Ukraine baseline models
fit_ukr_arima <- auto.arima(ukr_train_log)
fit_ukr_ets   <- ets(ukr_train_log)
fit_ukr_tbats <- tbats(ukr_train_log)

#Forecast 2022 war period baseline
h_ukr <- length(ukr_test_log)

fc_ukr_arima <- forecast(fit_ukr_arima, h = h_ukr)
fc_ukr_ets   <- forecast(fit_ukr_ets, h = h_ukr)
fc_ukr_tbats <- forecast(fit_ukr_tbats, h = h_ukr)

```

#ARIMA + War Shock
```{r}
pred_ukr_arima_war_log <- as.numeric(fc_ukr_arima$mean) + shock_gulf_arima
pred_ukr_arima_war <- exp(pred_ukr_arima_war_log)
```

#ETS + War shock
```{r}
pred_ukr_ets_war_log <- as.numeric(fc_ukr_ets$mean) + shock_gulf_ets
pred_ukr_ets_war <- exp(pred_ukr_ets_war_log)
```

#TBATS + War shock
```{r}
pred_ukr_tbats_war_log <- as.numeric(fc_ukr_tbats$mean) + shock_gulf_tbats
pred_ukr_tbats_war <- exp(pred_ukr_tbats_war_log)
```

#Compare forecasting data and real data
```{r}
actual_ukr <- as.numeric(ukr_test)
Arima_Scores<-accuracy(pred_ukr_arima_war, actual_ukr)
ETS_Scores<-accuracy(pred_ukr_ets_war, actual_ukr)
TBAT_Scores<-accuracy(pred_ukr_tbats_war, actual_ukr)

scores_cbind <- as.data.frame(rbind(Arima_Scores, ETS_Scores, TBAT_Scores))
row.names(scores_cbind) <-c("ARIMA","ETS","TBAT")
print(scores_cbind)
```
> We found that ARIMA is the best model.

#Use Ukraine War's shock to predict WTI price after 2026.4
```{r}
#Ukraine war shock 
shock_ukr_arima <- as.numeric(ukr_test_log - fc_ukr_arima$mean)

# check the shock values
print(shock_ukr_arima)

#Year 2026's WTI price
wti_train_2026 <- window(wti_ts, start = c(2022, 9), end = c(2026, 3)) #Use data after the Ukraine War
wti_train_2026_log <- log(wti_train_2026)

fit_final_arima <- auto.arima(wti_train_2026_log)
summary(fit_final_arima)

#Forecast baseline WTI after April 2026 
h_future <- length(shock_ukr_arima)

fc_2026_arima <- forecast(fit_final_arima, h = h_future)

# baseline forecast in log scale
baseline_2026_log <- as.numeric(fc_2026_arima$mean)

#Add 2022 Ukraine War shock trend to the 2026 baseline forecast
pred_2026_war_log <- baseline_2026_log + shock_ukr_arima
pred_2026_war <- exp(pred_2026_war_log)

#output
forecast_2026_results <- data.frame(
  Time = time(fc_2026_arima$mean),
  ARIMA_with_2022_WarShock = pred_2026_war
)

print(forecast_2026_results)
```

#Graph the forecast result
```{r}
# Filter observed data from 2021 to 2026.3
wti_plot_actual <- window(wti_ts, start = c(2021, 1), end = c(2026, 3))


#Turn forecast data into time series
wti_forecast_ts <- ts(
  pred_2026_war,
  start = c(2026, 4),
  frequency = 12
)

#Plot observed data
plot(
  wti_plot_actual,
  xlim = c(2021, 2026 + 10/12),
  ylim = range(c(wti_plot_actual, wti_forecast_ts)),
  col = "black",
  lwd = 2,
  ylab = "WTI Price",
  xlab = "Time",
  main = "WTI Price (2021–2026) with Forecast"
)+
  abline(v = 2026 + 3/12, col = "blue", lty = 3, lwd = 2)+
  lines(
  wti_forecast_ts,
  col = "red",
  lwd = 2,
  lty = 2 
)

legend(
  "topleft",
  legend = c("Actual WTI", "Forecast (ARIMA + War Shock)", "Forecast Start"),
  col = c("black", "red", "blue"),
  lty = c(1, 2, 3),
  lwd = 2,
  cex = 0.8
)
```

















