---
title: "Final report script"
authors: Trudy, Jenny, Janice, ZZ
date: "2026-04-04"
output:
  pdf_document: default
  html_document: default
---

Final Report

1. Introduction 

Oil prices are one of the most closely watched indicators in the global economy. (He et al., 2010) As the primary U.S. crude oil benchmark, West Texas Intermediate (WTI) spot prices impact a range of factors, including gasoline costs, inflation, supply chain dynamics, and energy policy. The volatility of oil prices is not solely driven by supply and demand fundamentals. Moreover, geopolitical shocks have historically triggered some of the most sudden and severe price disruptions in the market. From the Gulf War in 1990 to the Russia-Ukraine War in 2022 and the ongoing U.S.-Israeli strikes on Iran in 2026, each major geopolitical crisis has left a distinct and measurable imprint on oil price dynamics. 

2. Motivation 

The motivation for this study emerged directly from a real-time event. Upon returning from spring break in April 2026, WTI crude surged past $111 per barrel, which is the biggest single-day jump in six years(Kopack, 2026). It is the first time U.S. average gas prices have topped $4 per gallon since 2022(Kottasová and Isidore, 2026). The trigger was the U.S.-Israeli strikes on Iran beginning February 28, 2026, which effectively halted tanker traffic through the Strait of Hormuz. The IEA described it as the largest supply disruption in the history of the global oil market(Oil Market Report - March 2026 – Analysis, 2026). This prompted a central question: can historical patterns of geopolitical oil price shocks be used to forecast the trajectory of current and future crises? 

3. Relevance 

Understanding how geopolitical shocks transmit into oil price behavior has significant implications for energy markets, policymakers, investors, and consumers. Unlike structural supply-demand shifts, geopolitical disruptions are sudden and difficult to anticipate, yet their price impacts often follow recognizable patterns across events. By treating past wars and crises as informative precedents, time series models can provide a systematic, data-driven framework for anticipating price trajectories during periods of elevated geopolitical risk. This approach is particularly relevant in the current moment, where energy market volatility directly affects household costs, inflation expectations, and broader macroeconomic stability. 

4. Objectives 

This study has three primary objectives. First, we aim to establish a counterfactual baseline: what oil prices would have looked like in the absence of geopolitical disruption. We used ARIMA, ETS, TBATS, and ARIMA + Fourier models or training on pre-war historical WTI price data first. Second, we extract and quantify the "war shock," defined as the difference between actual oil prices and the counterfactual baseline during the Gulf War period, and transfer this shock pattern to simulate predicted oil price behavior during the Russia-Ukraine War. We then validate our approach by comparing simulated versus actual prices during 2022, finding ARIMA to be the best-performing model. Third, we apply the Russia-Ukraine War shock pattern to the current price level to forecast the oil price trajectory following the Iran/Hormuz crisis of 2026. Our dataset comprises monthly WTI spot prices, U.S. crude oil stocks, and U.S. field production from the EIA spanning 1986 to 2025. This period was selected for its comprehensive coverage of multiple geopolitical cycles and its relevance to the U.S. energy market context. 

5. Dataset information 

This analysis uses three monthly time-series datasets, which are Cushing, OK WTI Spot Price FOB, U.S. Ending Stocks of Crude Oil, and U.S. Field Production of Crude Oil. All of these data are downloaded from the Energy Information Administration (EIA). These three series are treated as the core inputs for the project. WTI spot price is the main outcome variable, while U.S. crude oil stocks and domestic field production are included as supply-side variables that may help explain oil price dynamics. The presentation also describes these variables as monthly measures of benchmark crude price, domestic supply buffer, and U.S. production conditions, respectively.  

Before getting into data analysis, we cleaned the data to make sure it is granular enough for future analysis. In the script, each dataset was converted into a data frame with two columns, date and numeric value, respectively. After importing the three files, we removed missing values with “na.omit()”, and observations were sorted chronologically using “arrange(Date)”. Next, the cleaned WTI series was converted into a time series object with “frequency = 12”. Because crude oil ending stocks are only available beginning in January 2005, all series used in the multivariable analysis were aligned to a common monthly window beginning in 2005 so that the variables shared a consistent time span. The script also constructed a binary war dummy for the Russia–Ukraine War period to test whether geopolitical shocks contributed significantly to price changes.

```{r setup}

library(readxl)
library(dplyr)
library(lubridate)
library(tidyr)
library(forecast)
library(tseries)
library(ggplot2)

```

We use three raw data series: oil price, U.S. oil stock, and U.S. domestic oil production. Oil price is the dependent variable of interest. U.S. oil stock and U.S. domestic oil production are included to account for supply-side conditions that may influence oil prices. We hypothesize that geopolitical events affect oil prices indirectly by changing U.S. oil inventories and, in turn, domestic oil production. These changes may then contribute to movements in oil prices.

## Data Input

```{r}
wti_raw <- read_excel("Data/Cushing, OK WTI Spot Price FOB (Dollars per Barrel).xls")
stock_raw <- read_excel("Data/U.S. Ending Stocks of Crude Oil (Thousand Barrels).xls")
prod_raw <- read_excel("Data/U.S. Field Production of Crude Oil (Thousand Barrels per Day).xls")
```

We are looking at a 10-year range because this would be long enough to observe how geopolitical events influenced oil price in the previous years. With that said, we will only select data

## Data Wrangling

```{r}
wti_df <- data.frame(
  Date  = as.Date(wti_raw[[1]], format = "%Y/%m/%d"),
  Price = as.numeric(wti_raw[[2]])
) %>%
  na.omit() %>%
  arrange(Date)

wti_ts <- ts(wti_df[,2], start = c(1986,1), frequency = 12)

wti_sub <- window(wti_ts, start = c(2005, 1), end = c(2025, 12))

# Plot WTI training series
plot(wti_sub, main = "WTI Price (2000-2025)", ylab = "Price", xlab = "Time")

# Stationarity check
adf.test(wti_sub)

fit_arima <- auto.arima(wti_sub)
fit_arima

```

# Involve fundamentals (production and inventores) and shock (war dummy) to test significance
We considered three candidate variables as potential drivers of WTI crude oil prices: U.S. domestic crude oil production, U.S. crude oil ending stocks, and a binary war dummy (coded as 1 during February–August 2022, coinciding with the onset and early escalation of the Russia–Ukraine War, and 0 otherwise). Since U.S. crude oil ending stocks data are available only from January 2005, all series were subset to the period beginning January 2005 to ensure a consistent sample across variables.

```{r include=FALSE}
#Stock
stock_df <- data.frame(
  Date  = as.Date(stock_raw[[1]], format = "%Y/%m/%d"),
  Price = as.numeric(stock_raw[[2]])
) %>%
  na.omit() %>%
  arrange(Date)

#Production
prod_df <- data.frame(
  Date  = as.Date(prod_raw[[1]], format = "%Y/%m/%d"),
  Price = as.numeric(prod_raw[[2]])
) %>%
  na.omit() %>%
  arrange(Date)

prod_ts  <- ts(prod_df[,2],  start = c(1920, 1), frequency = 12)
stock_ts <- ts(stock_df[,2], start = c(2005, 1), frequency = 12)

# Subset to same training period (2000–2025)
prod_sub  <- window(prod_ts,  start = c(2005,1), end = c(2025,12))
plot(prod_sub, main = "U.S Domestic Oil Production (2000-2025)", ylab = "Thousands Barrels per Day", xlab = "Time")

stock_sub <- window(stock_ts, start = c(2005,1), end = c(2025,12))
plot(stock_sub, main = "U.S Oil Stock (2000-2025)", ylab = "Thousands Barrels", xlab = "Time")

#War Dummy
time_index <- time(wti_sub)

war_dummy <- ifelse(
 time_index >= 2022 + 2/12 &
 time_index < 2022 + 6/12,
1,0)

#print(war_dummy) #check war dummy 
xreg_all <- cbind(prod_sub, stock_sub, war_dummy)

#print(xreg_all) #check all factors
fit_arimax <- auto.arima(wti_sub, xreg = xreg_all)
fit_arimax
```

To test the explanatory power of each variable within a time-series framework, we estimated an ARIMAX model incorporating all three variables as external regressors, which accounts for the serial correlation inherent in monthly oil price data by modeling the error term as an ARIMA process. The results indicated that domestic production and crude oil ending stocks were statistically insignificant, with near-zero coefficients relative to their standard errors. Both variables were therefore excluded from further analysis—domestic production on the basis of statistical insignificance. The war dummy, by contrast, returned a statistically significant positive coefficient of approximately $14 per barrel (t is about 3.44), confirming that geopolitical shocks constitute a meaningful and discrete disruption to oil price dynamics that warrants further investigation.

```{r echo=FALSE}
fit_arimax <- auto.arima(wti_sub, xreg = xreg_all)
fit_arimax
```
# Use Gulf War to train and test for Ukraine War
we adopt a historical analogy framework using the 1990–1991 Gulf War as a structural template. The underlying identification assumption is that major geopolitical oil supply shocks generate structurally similar patterns of price disturbance, allowing the shock profile of a historical conflict to serve as a proxy for that of a more recent one.

Four baseline forecasting models (ARIMA, ETS, TBATS, and ARIMA with Fourier) were estimated on WTI prices from January 1986 to June 1990, the pre-Gulf War period. The forecasts were generated over July 1990–January 1991 (seven months), representing the expected price trajectory in the absence of conflict. The war shock vector for each model was computed as the difference between actual and forecast prices, yielding a 7-month series that captures the dynamic shape of the geopolitical price disturbance. Results are reported in the table below.
```{r include=FALSE}
# Gulf War window
gulf_train <- window(wti_ts, start = c(1986, 1), end = c(1990, 6))
gulf_test  <- window(wti_ts, start = c(1990, 7), end = c(1991, 1))

#log transform
gulf_train_log <- log(gulf_train)
gulf_test_log  <- log(gulf_test)

#Gulf War baseline model
fit_gulf_arima <- auto.arima(gulf_train_log)
fit_gulf_ets   <- ets(gulf_train_log)
fit_gulf_tbats <- tbats(gulf_train_log)

# ARIMA + Fourier
best_aicc_gulf <- Inf
best_k_gulf <- NA
fit_gulf_fourier <- NULL

for(k in 1:3){
  fit_try <- auto.arima(
    gulf_train_log,
    xreg = fourier(gulf_train_log, K = k),
    seasonal = FALSE
  )
  
  if(fit_try$aicc < best_aicc_gulf){
    best_aicc_gulf <- fit_try$aicc
    best_k_gulf <- k
    fit_gulf_fourier <- fit_try
  }
}

#Forecast Gulf War based on baseline
h_gulf <- length(gulf_test_log)

fc_gulf_arima <- forecast(fit_gulf_arima, h = h_gulf)
fc_gulf_ets   <- forecast(fit_gulf_ets, h = h_gulf)
fc_gulf_tbats <- forecast(fit_gulf_tbats, h = h_gulf)

fc_gulf_fourier <- forecast(
  fit_gulf_fourier,
  xreg = fourier(gulf_train_log, K = best_k_gulf, h = h_gulf),
  h = h_gulf
)

# War shock effects
shock_gulf_arima <- as.numeric(gulf_test_log - fc_gulf_arima$mean)
shock_gulf_ets   <- as.numeric(gulf_test_log - fc_gulf_ets$mean)
shock_gulf_tbats <- as.numeric(gulf_test_log - fc_gulf_tbats$mean)
shock_gulf_fourier <- as.numeric(gulf_test_log - fc_gulf_fourier$mean)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
shock_table <- data.frame(
  Model = c("ARIMA", "ETS", "TBATS", "ARIMA+Fourier"),
  Month_1 = c(shock_gulf_arima[1], shock_gulf_ets[1], shock_gulf_tbats[1], shock_gulf_fourier[1]),
  Month_2 = c(shock_gulf_arima[2], shock_gulf_ets[2], shock_gulf_tbats[2], shock_gulf_fourier[2]),
  Month_3 = c(shock_gulf_arima[3], shock_gulf_ets[3], shock_gulf_tbats[3], shock_gulf_fourier[3]),
  Month_4 = c(shock_gulf_arima[4], shock_gulf_ets[4], shock_gulf_tbats[4], shock_gulf_fourier[4]),
  Month_5 = c(shock_gulf_arima[5], shock_gulf_ets[5], shock_gulf_tbats[5], shock_gulf_fourier[5]),
  Month_6 = c(shock_gulf_arima[6], shock_gulf_ets[6], shock_gulf_tbats[6], shock_gulf_fourier[6]),
  Month_7 = c(shock_gulf_arima[7], shock_gulf_ets[7], shock_gulf_tbats[7], shock_gulf_fourier[7])
)

colnames(shock_table) <- c("Model", "Month 1", "Month 2", "Month 3", "Month 4", "Month 5", "Month 6", "Month 7")

knitr::kable(shock_table, digits = 4, caption = "Gulf War Shock Vectors by Model (Log Scale)")
```
# Russia-Ukraine War's effect forecast

The same four models were re-estimated on WTI prices from January 2018 to January 2022, and forecasts were generated for February–August 2022. The Gulf War shock vector was added to each model's predicted baseline to produce a war-adjusted price forecast for the Ukraine War period. Forecast accuracy was evaluated against realized WTI prices using RMSE, MAE, and MAPE.

```{r include=FALSE}
# Ukraine window（same length as Gulf War）
ukr_train <- window(wti_ts, start = c(2018, 1), end = c(2022, 1))
ukr_test  <- window(wti_ts, start = c(2022, 2), end = c(2022, 8))  # 7 months

#log transform
ukr_train_log <- log(ukr_train) 
ukr_test_log  <- log(ukr_test)

#Ukraine baseline models
fit_ukr_arima <- auto.arima(ukr_train_log)
fit_ukr_ets   <- ets(ukr_train_log)
fit_ukr_tbats <- tbats(ukr_train_log)

# ARIMA + Fourier
best_aicc_ukr <- Inf
best_k_ukr <- NA
fit_ukr_fourier <- NULL

for(k in 1:3){
  fit_try <- auto.arima(
    ukr_train_log,
    xreg = fourier(ukr_train_log, K = k),
    seasonal = FALSE
  )
  
  if(fit_try$aicc < best_aicc_ukr){
    best_aicc_ukr <- fit_try$aicc
    best_k_ukr <- k
    fit_ukr_fourier <- fit_try
  }
}

#Forecast 2022 war period baseline
h_ukr <- length(ukr_test_log)

fc_ukr_arima <- forecast(fit_ukr_arima, h = h_ukr)
fc_ukr_ets   <- forecast(fit_ukr_ets, h = h_ukr)
fc_ukr_tbats <- forecast(fit_ukr_tbats, h = h_ukr)

fc_ukr_fourier <- forecast(
  fit_ukr_fourier,
  xreg = fourier(ukr_train_log, K = best_k_ukr, h = h_ukr),
  h = h_ukr
)

```


```{r include=FALSE}
#ARIMA + War Shock
pred_ukr_arima_war_log <- as.numeric(fc_ukr_arima$mean) + shock_gulf_arima
pred_ukr_arima_war <- exp(pred_ukr_arima_war_log)
```


```{r include=FALSE}
#ETS + War shock
pred_ukr_ets_war_log <- as.numeric(fc_ukr_ets$mean) + shock_gulf_ets
pred_ukr_ets_war <- exp(pred_ukr_ets_war_log)
```


```{r include=FALSE}
#TBATS + War shock
pred_ukr_tbats_war_log <- as.numeric(fc_ukr_tbats$mean) + shock_gulf_tbats
pred_ukr_tbats_war <- exp(pred_ukr_tbats_war_log)
```


```{r include=FALSE}
# ARIMA + Fourier + War shock
pred_ukr_fourier_war_log <- as.numeric(fc_ukr_fourier$mean) + shock_gulf_fourier
pred_ukr_fourier_war <- exp(pred_ukr_fourier_war_log)
```

The results of predicted WTI prices during the Russia-Ukraine War period are listed below. 
```{r echo=FALSE, message=FALSE, warning=FALSE}
shock_table <- data.frame(
  Model = c("ARIMA", "ETS", "TBATS", "ARIMA+Fourier"),
  Month_1 = c(pred_ukr_arima_war[1], pred_ukr_ets_war[1], pred_ukr_tbats_war[1], pred_ukr_fourier_war[1]),
  Month_2 = c(pred_ukr_arima_war[2], pred_ukr_ets_war[2], pred_ukr_tbats_war[2], pred_ukr_fourier_war[2]),
  Month_3 = c(pred_ukr_arima_war[3], pred_ukr_ets_war[3], pred_ukr_tbats_war[3], pred_ukr_fourier_war[3]),
  Month_4 = c(pred_ukr_arima_war[4], pred_ukr_ets_war[4], pred_ukr_tbats_war[4], pred_ukr_fourier_war[4]),
  Month_5 = c(pred_ukr_arima_war[5], pred_ukr_ets_war[5], pred_ukr_tbats_war[5], pred_ukr_fourier_war[5]),
  Month_6 = c(pred_ukr_arima_war[6], pred_ukr_ets_war[6], pred_ukr_tbats_war[6], pred_ukr_fourier_war[6]),
  Month_7 = c(pred_ukr_arima_war[7], pred_ukr_ets_war[7], pred_ukr_tbats_war[7], pred_ukr_fourier_war[7])
)

colnames(shock_table) <- c("Model", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug")

knitr::kable(shock_table, digits = 4, caption = "Forecast WTI Price during Russia-Ukraine War Period")
```

We then compared the predicted results with the real-world data, using the code accuracy() to find the model that gave the best value. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Compare forecasting data and real data
actual_ukr <- as.numeric(ukr_test)
Arima_Scores<-accuracy(pred_ukr_arima_war, actual_ukr)
ETS_Scores<-accuracy(pred_ukr_ets_war, actual_ukr)
TBAT_Scores<-accuracy(pred_ukr_tbats_war, actual_ukr)
Fourier_Scores <- accuracy(pred_ukr_fourier_war, actual_ukr)

scores_cbind <- as.data.frame(rbind(Arima_Scores, ETS_Scores, TBAT_Scores, Fourier_Scores))
row.names(scores_cbind) <-c("ARIMA","ETS","TBAT","Fourier")
print(scores_cbind)
```

The ARIMA+Fourier specification achieved the lowest error across all three metrics and was selected as the final model for subsequent forecasting.

# Use Russia-Ukraine War's shock to predict WTI price after 2026.4

To forecast the potential WTI price impact of the current Middle East conflict, a final ARIMA+Fourier baseline model was estimated on WTI prices from September 2022 to March 2026—the post-Ukraine War period. A 7-month forecast was generated from April 2026 onward, and the Ukraine War shock vector derived from the ARIMA+Fourier model was added to this baseline to produce the final war-adjusted price forecast. This assumes that the current conflict generates a price disturbance pattern analogous in structure to that observed during the Russia–Ukraine War.

```{r include=FALSE}
# Ukraine war shock from Fourier model
shock_ukr_fourier <- as.numeric(ukr_test_log - fc_ukr_fourier$mean)

# check the shock values
print(shock_ukr_fourier)

# Year 2026's WTI price
wti_train_2026 <- window(wti_ts, start = c(2022, 9), end = c(2026, 3))  # Use data after the Ukraine War
wti_train_2026_log <- log(wti_train_2026)

# Final ARIMA + Fourier model
best_aicc_2026 <- Inf
best_k_2026 <- NA
fit_final_fourier <- NULL

for(k in 1:3){
  fit_try <- auto.arima(
    wti_train_2026_log,
    xreg = fourier(wti_train_2026_log, K = k),
    seasonal = FALSE
  )
  
  if(fit_try$aicc < best_aicc_2026){
    best_aicc_2026 <- fit_try$aicc
    best_k_2026 <- k
    fit_final_fourier <- fit_try
  }
}

summary(fit_final_fourier)

# Forecast baseline WTI after April 2026
h_future <- length(shock_ukr_fourier)

fc_2026_fourier <- forecast(
  fit_final_fourier,
  xreg = fourier(wti_train_2026_log, K = best_k_2026, h = h_future),
  h = h_future
)

# baseline forecast in log scale
baseline_2026_log <- as.numeric(fc_2026_fourier$mean)

# Add 2022 Ukraine War shock trend to the 2026 baseline forecast
pred_2026_war_log <- baseline_2026_log + shock_ukr_fourier
pred_2026_war <- exp(pred_2026_war_log)

# output
forecast_2026_results <- data.frame(
  Time = time(fc_2026_fourier$mean),
  Fourier_with_2022_WarShock = pred_2026_war
)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
forecast_display <- data.frame(
  Date = c("Apr 2026", "May 2026", "Jun 2026", "Jul 2026", "Aug 2026", "Sep 2026", "Oct 2026"),
  Predicted_WTI_Price = round(forecast_2026_results$Fourier_with_2022_WarShock, 2)
)

colnames(forecast_display) <- c("Date", "Predicted WTI Price ($/barrel)")

knitr::kable(forecast_display, caption = "Predicted WTI Crude Oil Prices (April–October 2026)")
```

# Graph the forecast result
```{r echo=FALSE}
# Filter observed data from 2021 to 2026.3
wti_plot_actual <- window(wti_ts, start = c(2021, 1), end = c(2026, 3))

# Turn forecast data into time series
wti_forecast_ts <- ts(
  pred_2026_war,
  start = c(2026, 4),
  frequency = 12
)

# Plot observed data
plot(
  wti_plot_actual,
  xlim = c(2021, 2026 + 10/12),
  ylim = range(c(wti_plot_actual, wti_forecast_ts)),
  col = "black",
  lwd = 2,
  ylab = "WTI Price",
  xlab = "Time",
  main = "WTI Price (2021–2026) with Forecast"
)

# Add forecast start line
abline(v = 2026 + 3/12, col = "blue", lty = 3, lwd = 2)

# Add forecast line
lines(
  wti_forecast_ts,
  col = "red",
  lwd = 2,
  lty = 2
)

# Add legend
legend(
  "topleft",
  legend = c("Actual WTI", "Forecast (Fourier + War Shock)", "Forecast Start"),
  col = c("black", "red", "blue"),
  lty = c(1, 2, 3),
  lwd = 2,
  cex = 0.8
)
```

6. Summary and Conclusions 

This project analyzed monthly WTI crude oil prices using several time series forecasting models, including ARIMA, ETS, TBATS, and ARIMA with Fourier terms, with a particular focus on how geopolitical shocks may affect oil price movements. The results show that ARIMA-based models produced the strongest forecasting performance among the models tested, suggesting that WTI prices are better explained by persistence, trend changes, and short-term shocks than by stable seasonal patterns. ETS and TBATS were less accurate, indicating that seasonality alone is not sufficient to capture the behavior of oil prices. 

Adding Fourier terms only provides modest improvement, which further supports the conclusion that recurring seasonal cycles are weaker than external disruptions in this market. In contrast, geopolitical events such as wars or regional conflicts can generate sudden and meaningful price deviations. Incorporating war-shock periods into the analysis helped demonstrate that external events are a major source of forecast uncertainty and should be treated explicitly rather than assumed to be random noise. 

The 2026 forecast suggests that oil prices may remain relatively elevated but gradually stabilize under normal conditions. However, if geopolitical tensions intensify, prices could rise above the baseline forecast path. This highlights the importance of interpreting forecasts as scenarios rather than fixed predictions, especially in commodity markets that are highly sensitive to political and supply-side developments. 

Overall, the findings suggest that ARIMA models provide a strong baseline framework for forecasting oil prices, but their predictive power improves when external shocks are incorporated more clearly. Future improvements could include using ARIMAX models with variables such as interest rates, GDP, and other quantifiable economic factors. Moreover, it's also worthwhile to apply regime-switching models that better capture structural breaks. Expanding the use of scenario analysis for geopolitical risk would also strengthen the practical value of the model. 















